---
title: "Slower recovery of soil fungal communities: insights from >20 years of tropical forest restoration in rubber monocultures"
format: html
editor: visual
---

# Load required packages

```{r message=FALSE, warning=FALSE}
library(patchwork); library(iCAMP); library(gridExtra); library(emmeans); library(tidyr); library(multcomp); library(TH.data); library(MASS); library(survival); library(mvtnorm); library(lmerTest); library(lme4); library(Matrix); library(combinat); library(ggpubr); library(ggplot2); library(dplyr); library(spdep); library(sf); library(spData); library(picante); library(nlme); library(vegan); library(permute); library(ape); library(gridExtra); library(tidyverse); library(knitr); library(vegan)

set.seed(1234)
```

# Import data

```{r}

# Raw data
OTU_table_total_fungi_unrarefied <- read.csv("data/total_fungi_unrarefied.csv", row.names = 1, header = TRUE)
OTU_table_AM_fungi_unrarefied <- read.csv("data/am_fungi_unrarefied.csv", row.names = 1, header = TRUE)

# Transform raw data to relative abundance
OTU_table_total_fungi_t <- t(OTU_table_total_fungi_unrarefied)
OTU_table_AM_fungi_t <- t(OTU_table_AM_fungi_unrarefied)

# Rarefied data
OTU_table_total_fungi <- read.csv("data/total_fungi_rarefied.csv", row.names = 1, header = TRUE)
OTU_table_AM_fungi <- read.csv("data/am_fungi_rarefied.csv", row.names = 1, header = TRUE)

# Transform rarefied data to relative abundance
total_rel_abundance <- sweep(OTU_table_total_fungi,2,colSums(OTU_table_total_fungi),FUN = "/")
AM_rel_abundance_AM <- sweep(OTU_table_AM_fungi,2,colSums(OTU_table_AM_fungi),FUN = "/")

# Import phylogenetic tree
total_tree <- read.tree("data/total_fungi_tree.nwk")
AM_tree <- read.tree("data/am_fungi_tree.nwk")

# Import environmental variable
env_var <- read.csv("data/env.csv", row.names = 1, header = TRUE)
location <- read.csv("data/location.csv", row.names = 1, header = TRUE)
```

# Data overview

## Rarefaction curves

```{r}
par(mfrow = c(1, 2),
    mar = c(4.2, 4.2, 2, 1),
    oma = c(1, 1, 1, 1),
    mgp = c(1.8, 0.6, 0),
    cex.lab = 0.8,
    cex.axis = 0.7,
    cex.main = 0.8,
    font.main = 1)

forest_colors <- c(RM = "#F8766D", NR = "#619CFF", RP = "cyan3", MF = "#00BA38")
curve_colors_total <- forest_colors[rep(c("RM", "NR", "RP", "MF"), each = 8)]
curve_colors_AM    <- forest_colors[rep(c("RM", "NR", "RP", "MF"), each = 8)]

rarecurve(OTU_table_total_fungi_t, step = 500, col = curve_colors_total,
          cex = 0.3, label = FALSE,
          xlab = "Sequencing depth",
          ylab = "Number of total fungal OTUs",
          main = "Total fungi")
legend(55000, 360, legend = names(forest_colors), col = forest_colors,
       lwd = 1, cex = 0.5, bty = "n")

rarecurve(OTU_table_AM_fungi_t, step = 500, col = curve_colors_AM,
          cex = 0.3, label = FALSE,
          xlab = "Sequencing depth",
          ylab = "Number of AM fungal VTs",
          main = "AM fungi")
legend(40000, 18, legend = names(forest_colors), col = forest_colors,
       lwd = 1, cex = 0.5, bty = "n")
```

## Relative abundance total fungi

```{r}
Relative_abundance_phylum <- read.csv("data/total_fungi_phylum.csv",header = TRUE)
Relative_abundance_phylum <- Relative_abundance_phylum %>% mutate(season=factor(season, levels=c("Wet","Dry")), forest=factor(forest, levels=c("RM","NR","RP","MF")), phylum=factor(phylum, levels=c("Unidentified","Ascomycota","Basidiomycota","Mortierellomycota","Mucoromycota","Basal lineage")))
manual_colors_functional <- c("Unidentified"="grey50","Ascomycota"="#006837","Basidiomycota"="#1a9850","Mortierellomycota"="#66bd63","Mucoromycota"="#a6d96a","Basal lineage"="#d9f0a3")
phylum_abundance <- Relative_abundance_phylum %>%group_by(phylum) %>%summarise(total_abundance = sum(abundance, na.rm = TRUE),.groups = "drop")
Total_fungi_family_pie_chart <- ggplot(phylum_abundance, aes(x="", y=total_abundance, fill=phylum)) + geom_bar(stat="identity", linewidth=1, alpha=0.6) + coord_polar(theta="y") + scale_fill_manual(values=manual_colors_functional) + theme_void() + labs(title="Total fungi", fill="Phylum") + theme(plot.background=element_rect(fill="white", color=NA, size=0.3), panel.background=element_rect(fill="white"), plot.title=element_text(hjust=0.5, size=10), legend.position="right", legend.title=element_text(size=9), legend.text=element_text(size=8), legend.key.size=unit(0.3,"cm"), aspect.ratio=1)
Total_fungi_family_pie_chart
```

## Relative abundance am fungi

```{r}
Relative_abundance_AM_family <- read.csv("data/am_fungi_family.csv", header = TRUE)
Relative_abundance_AM_family$season<-as.factor(Relative_abundance_AM_family$season)
Relative_abundance_AM_family$forest<-as.factor(Relative_abundance_AM_family$forest)
Relative_abundance_AM_family$family<-as.factor(Relative_abundance_AM_family$family)
Relative_abundance_AM_family$family <- factor(Relative_abundance_AM_family$family, levels = c("Glomeraceae","Acaulosporaceae","Claroideoglomeraceae","Archaeosporaceae","Gigasporaceae","Paraglomeraceae","Diversisporaceae"))
Relative_abundance_AM_family$forest<-factor(Relative_abundance_AM_family$forest,levels=c("RM", "NF", "AF", "SF"))
Relative_abundance_AM_family$season<-factor(Relative_abundance_AM_family$season,levels=c("Wet", "Dry"))
Relative_abundance_AM_family_scaled <- Relative_abundance_AM_family %>% group_by(forest, season) %>% mutate(Relative_abundance_percent = abundance / sum(abundance) * 100)
manual_colors_family <- c("Glomeraceae"="#0B1D51","Acaulosporaceae"="#1034A6","Claroideoglomeraceae"="#0066CC","Archaeosporaceae"="#1E90FF","Gigasporaceae"="#3399FF","Paraglomeraceae"="#66B2FF","Diversisporaceae"="#99CCFF","Others"="#CFE7FF")
AM_fungi_family_pie_chart <- ggplot(Relative_abundance_AM_family_scaled, aes(x="", y=Relative_abundance_percent, fill=family)) + geom_bar(stat="identity", linewidth=1, alpha=0.6) + coord_polar(theta="y") + scale_fill_manual(values=manual_colors_family) + labs(title="AM fungi", fill="Family") + theme_void() + theme(plot.background=element_rect(fill="white", color=NA, size=0.3), panel.background=element_rect(fill="white"), plot.title=element_text(hjust=0.5, size=10), legend.position="right", legend.title=element_text(size=8.9), legend.text=element_text(size=7.9), legend.key.size=unit(0.3,"cm"), aspect.ratio=1, plot.margin=margin(10,10,10,10,unit="pt"))
AM_fungi_family_pie_chart
```

# Alpha diversity

## Total fungi

```{r}
richness_total <- colSums(total_rel_abundance > 0)
shannon_total <- diversity(t(total_rel_abundance), index = "shannon")
OTU_pa_total <- t(total_rel_abundance > 0)
PD_total <- pd(OTU_pa_total, total_tree, include.root = FALSE)
alpha_total <- data.frame(Plot_ID=rownames(PD_total), Richness=richness_total[rownames(PD_total)], Shannon=shannon_total[rownames(PD_total)], PD=PD_total$PD)
alpha_total <- cbind(alpha_total, env_var, location)
alpha_total$Plot_ID <- as.character(alpha_total$Plot_ID)
alpha_total$Plot_ID <- sub("^XS", "SB", alpha_total$Plot_ID)
alpha_total$restoration <- factor(ifelse(alpha_total$forest_types=="RM","Monoculture", ifelse(alpha_total$forest_types %in% c("NR","RP"),"Restored","Reference")), levels=c("Monoculture","Restored","Reference"))
```

## AM fungi

```{r}
richness_AM <- colSums(AM_rel_abundance_AM > 0)
shannon_AM <- diversity(t(AM_rel_abundance_AM), index = "shannon")
OTU_pa_AM <- t(AM_rel_abundance_AM > 0)
PD_AM <- pd(OTU_pa_AM, AM_tree, include.root = FALSE)
alpha_AM <- data.frame(Plot_ID=rownames(PD_AM), Richness=richness_AM[rownames(PD_AM)], Shannon=shannon_AM[rownames(PD_AM)], PD=PD_AM$PD)
alpha_AM <- cbind(alpha_AM, env_var, location)
alpha_AM$Plot_ID <- as.character(alpha_AM$Plot_ID)
alpha_AM$Plot_ID <- sub("^XS", "SB", alpha_AM$Plot_ID)
alpha_AM$restoration <- factor(ifelse(alpha_AM$forest_types=="RM","Monoculture", ifelse(alpha_AM$forest_types %in% c("NR","RP"),"Restored","Reference")), levels=c("Monoculture","Restored","Reference"))
```

## Figure 1

```{r}
alpha_long <- alpha_total %>% pivot_longer(cols = c(Richness, Shannon, PD), names_to = "Metric", values_to = "Value")
alpha_long$forest_types <- factor (alpha_long$forest_types, levels = c("RM", "NR", "RP", "MF"))
alpha_long$Metric <- factor(alpha_long$Metric, levels = c("Richness", "Shannon", "PD"), labels = c("Observed richness", "Shannon diversity", "Phylogenetic diversity"))
common_colors <- c("RM"= "cyan4", "NR" = "chartreuse3", "RP"= "#7db53eff", "MF" = "forestgreen")
common_shapes <- c("Dry" = 16, "Wet" = 17)
Figure_1_1 <- ggplot(alpha_long, aes(x = forest_types, y = Value, fill = forest_types)) + geom_boxplot(width = 0.5, outlier.shape = NA) + geom_jitter(aes(shape = season), width = 0.05, size = 1.5, alpha = 0.7) + geom_vline(xintercept = 1.6, color = "grey50", linetype = "dashed") + geom_vline(xintercept = 3.5, color = "grey50", linetype = "dashed") + scale_fill_manual(values = common_colors) + facet_wrap(~ Metric, scales = "free_y") + theme_pubr(border = TRUE, legend = "none") + labs(x = "Forest type", y = "Alpha diversity (Total fungi)", shape = "Season") + theme(strip.background = element_rect(fill = "grey20"), strip.text = element_text(size = 12, face = "plain", color = "white")) + annotate("text", x = 1, y = Inf, label = "Monoculture", vjust = 1.8, size = 3.2, fontface = "plain") + annotate("text", x = 2.5, y = Inf, label = "Restored forests", vjust = 1.8, size = 3.2, fontface = "plain") + annotate("text", x = 4, y = Inf, label = "Rainforest", vjust = 1.8, size = 3.2, fontface = "plain") + scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
Figure_1_1
```

```{r}
alpha_AM_long <- alpha_AM %>% pivot_longer(cols = c(Richness, Shannon, PD), names_to = "Metric", values_to = "Value")
alpha_AM_long$forest_types <- factor(alpha_AM_long$forest_types, levels = c("RM", "NR", "RP", "MF"))
alpha_AM_long$Metric <- factor(alpha_AM_long$Metric, levels = c("Richness", "Shannon", "PD"), labels = c("Observed richness", "Shannon diversity", "Phylogenetic diversity"))
Figure_1_2 <- ggplot(alpha_AM_long, aes(x = forest_types, y = Value, fill = forest_types)) + geom_boxplot(width = 0.5, outlier.shape = NA) + geom_jitter(aes(shape = season), width = 0.05, size = 1.5, alpha = 0.7) + geom_vline(xintercept = 1.6, color = "grey50", linetype = "dashed") + geom_vline(xintercept = 3.5, color = "grey50", linetype = "dashed") + scale_fill_manual(values = common_colors) + facet_wrap(~ Metric, scales = "free_y") + theme_pubr(border = TRUE, legend = "none") + labs(x = "Forest type", y = "Alpha diversity (AM fungi)", shape = "Season") + theme(strip.background = element_rect(fill = "grey20"), strip.text = element_text(size = 12, face = "plain", color = "white")) + annotate("text", x = 1, y = Inf, label = "Monoculture", vjust = 1.8, size = 3.2, fontface = "plain") + annotate("text", x = 2.5, y = Inf, label = "Restored forests", vjust = 1.8, size = 3.2, fontface = "plain") + annotate("text", x = 4, y = Inf, label = "Rainforest", vjust = 1.8, size = 3.2, fontface = "plain") + scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
Figure_1_2
```

## Significance test

### Test spatial autocorrelation

```{r}
set.seed(123) 

location_jittered <- location %>% 
  mutate(latitude_jitter  = jitter(Latitude,  amount = 0.00001),longitude_jitter = jitter(Longitude, amount = 0.00001))

# Spatial weights (based ONLY on coordinates)
coords <- cbind(location_jittered$longitude_jitter,location_jittered$latitude_jitter)
nb <- knn2nb(knearneigh(coords, k = 8))
lw <- nb2listw(nb, style = "W")

run_moran_test <- function(formula, data, listw) {
  mod <- lm(formula, data = data)
  res <- resid(mod)
  moran.test(res, listw)
}

metrics <- list(Richness = Richness ~ forest_types * season,Shannon = Shannon ~ forest_types * season, PD = PD ~ forest_types * season)
alpha_total_spatial <- alpha_total[,c("Richness", "Shannon", "PD", "season", "forest_types")]
moran_results_total <- lapply(names(metrics), function(name) {
  cat("\n### Moran's I – Total fungi:", name, "###\n")
  result <- run_moran_test(metrics[[name]], alpha_total_spatial, lw)
  print(result)
  result
})

names(moran_results_total) <- names(metrics)
alpha_AM_spatial <- alpha_AM[, c("Richness", "Shannon", "PD", "season", "forest_types")]
moran_results_AM <- lapply(names(metrics), function(name) {
  cat("\n### Moran's I – AM fungi:", name, "###\n")
  result <- run_moran_test(metrics[[name]], alpha_AM_spatial, lw)
  print(result)
  result
})

names(moran_results_AM) <- names(metrics)
extract_moran <- function(moran_list) {
  data.frame(
    Metric = names(moran_list),
    Moran_I = sapply(moran_list, function(x) x$estimate[1]),
    Expected_I = sapply(moran_list, function(x) x$estimate[2]),
    Variance = sapply(moran_list, function(x) x$estimate[3]),
    Z_score = sapply(moran_list, function(x) x$statistic),
    P_value = sapply(moran_list, function(x) x$p.value),
    row.names = NULL
  )
}

moran_total_table <- extract_moran(moran_results_total)
kable(moran_total_table, digits = 4, caption = "Moran’s I test for spatial autocorrelation in residuals of total fungal diversity models")
moran_AM_table <- extract_moran(moran_results_AM)
kable(moran_AM_table, digits = 4, caption = "Moran’s I test for spatial autocorrelation in residuals of AM fungal diversity models")
```

### Effect of restoration in alpha diversity

#### Richness

##### Total fungi

Does restoration improve richness?

```{r}
alpha_total$restoration <- relevel(alpha_total$restoration, ref = "Monoculture")
shapiro.test(alpha_total$Richness)
mod_restoration <- lmer(Richness ~ restoration + season + (1 | Plot_ID), data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Does restored forest level rainforest condition?

```{r}
alpha_total$restoration <- relevel(alpha_total$restoration, ref = "Reference")
mod_restoration <- lmer(Richness ~ restoration + season + (1 | Plot_ID), data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Are restoration strategies different?

```{r}
alpha_total$forest_types <- factor(alpha_total$forest_types)
alpha_total$forest_types <- relevel(alpha_total$forest_types, ref = "NR")
mod_restoration <- lmer(Richness ~ forest_types + season + (1 | Plot_ID),data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

##### AM fungi

Does restoration improve richness?

```{r}
alpha_AM$restoration <- relevel(alpha_AM$restoration, ref = "Monoculture")
shapiro.test(alpha_AM$Richness)
mod_restoration <- lmer(Richness ~ restoration + season + (1 | Plot_ID),data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Does restored forest level rainforest condition?

```{r}
alpha_AM$restoration <- relevel(alpha_AM$restoration, ref = "Reference")
mod_restoration <- lmer(Richness ~ restoration + season + (1 | Plot_ID),data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Are restoration strategies different?

```{r}
alpha_AM$forest_types <- factor(alpha_AM$forest_types)
alpha_AM$forest_types <- relevel(alpha_AM$forest_types, ref = "NR")
mod_restoration <- lmer(Richness ~ forest_types + season + (1 | Plot_ID),data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

#### Shannon

##### Total fungi

Does restoration improve Shannon diversity?

```{r}
alpha_total$restoration <- relevel(alpha_total$restoration, ref = "Monoculture")
shapiro.test(alpha_total$Shannon)
mod_restoration <- lmer(Shannon ~ restoration + season + (1 | Plot_ID),data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Does restoration improve Shannon diversity?

```{r}
alpha_total$restoration <- relevel(alpha_total$restoration, ref = "Reference")
mod_restoration <- lmer(Shannon ~ restoration + season + (1 | Plot_ID),data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Are restoration strategies different? (forest types)

```{r}
alpha_total$forest_types <- factor(alpha_total$forest_types)
alpha_total$forest_types <- relevel(alpha_total$forest_types, ref = "NR")
mod_restoration <- lmer(Shannon ~ forest_types + season + (1 | Plot_ID),data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

##### AM fungi

Does restoration improve Shannon diversity?

```{r}
alpha_AM$restoration <- relevel(alpha_AM$restoration, ref = "Monoculture")
shapiro.test(alpha_AM$Shannon)
mod_restoration <- lmer(Shannon ~ restoration + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Does restored forest level rainforest condition?

```{r}
alpha_AM$restoration <- relevel(alpha_AM$restoration, ref = "Reference")
mod_restoration <- lmer(Shannon ~ restoration + season + (1 | Plot_ID),data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Are restoration strategies different? (forest types)

```{r}
alpha_AM$forest_types <- factor(alpha_AM$forest_types)
alpha_AM$forest_types <- relevel(alpha_AM$forest_types, ref = "NR")
mod_restoration <- lmer(Shannon ~ forest_types + season + (1 | Plot_ID),data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

#### Phylogenetic diversity

##### Total fungi

Does restoration improve Phylogenetic Diversity (PD)?

```{r}
alpha_total$restoration <- relevel(alpha_total$restoration, ref = "Monoculture")
shapiro.test(alpha_total$PD)
mod_restoration <- lmer(PD ~ restoration + season + (1 | Plot_ID),data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Does restored forest level rainforest condition?

```{r}
alpha_total$restoration <- relevel(alpha_total$restoration, ref = "Reference")
mod_restoration <- lmer(PD ~ restoration + season + (1 | Plot_ID),data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Are restoration strategies different? (forest types)

```{r}
alpha_total$forest_types <- factor(alpha_total$forest_types)
alpha_total$forest_types <- relevel(alpha_total$forest_types, ref = "NR")
mod_restoration <- lmer(PD ~ forest_types + season + (1 | Plot_ID),data = alpha_total)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

##### AM fungi

Does restoration improve Phylogenetic Diversity (PD)?

```{r}
alpha_AM$restoration <- relevel(alpha_AM$restoration, ref = "Monoculture")
shapiro.test(alpha_AM$PD)
mod_restoration <- lmer(PD ~ restoration + season + (1 | Plot_ID),data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Does restored forest level rainforest condition?

```{r}
alpha_AM$restoration <- relevel(alpha_AM$restoration, ref = "Reference")
mod_restoration <- lmer(PD ~ restoration + season + (1 | Plot_ID),data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

Are restoration strategies different? (forest types)

```{r}
alpha_AM$forest_types <- factor(alpha_AM$forest_types)
alpha_AM$forest_types <- relevel(alpha_AM$forest_types, ref = "NR")
mod_restoration <- lmer(PD ~ forest_types + season + (1 | Plot_ID),data = alpha_AM)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

# Community composition

## Total fungi

### Taxonomic composition

```{r}
total_rel_abundance_t <- t(total_rel_abundance)
bray_dist_total <- vegdist(total_rel_abundance_t, method = "bray")
bray_mat <- as.matrix(bray_dist_total)
total_var <- sum(bray_mat^2) / nrow(bray_mat)
total_var

# PCoA
bray_pcoa_total <- cmdscale(bray_dist_total, eig = TRUE, k = 2)

# PCoA data frame
pcoa_bray_df <- data.frame(Sample = rownames(bray_pcoa_total$points), PCoA1 = bray_pcoa_total$points[, 1], PCoA2 = bray_pcoa_total$points[, 2], forest_type = env_var$forest_type, season = env_var$season, facet = "Total fungi taxonomic composition")

# Variance explained
var_exp_bray1 <- round(bray_pcoa_total$eig[1] / sum(bray_pcoa_total$eig) * 100, 1)
var_exp_bray2 <- round(bray_pcoa_total$eig[2] / sum(bray_pcoa_total$eig) * 100, 1)

# Factor levels
pcoa_bray_df$forest_type <- factor(pcoa_bray_df$forest_type, levels = c("RM", "NR", "RP", "MF"))
pcoa_bray_df$season <- factor(pcoa_bray_df$season, levels = c("Dry", "Wet"))

# Define common color and shape scales
common_colors <- c("RM"= "cyan4", "NR" = "chartreuse3", "RP"= "#7db53eff", "MF" = "forestgreen")
common_shapes <- c("Dry" = 16, "Wet" = 17)

species_composition_total <- ggplot(pcoa_bray_df, aes(x=PCoA1, y=PCoA2, colour=forest_type, shape=season)) + geom_point(size=5, stroke=1) + geom_vline(xintercept=0, linetype="dashed", color="grey50") + geom_hline(yintercept=0, linetype="dashed", color="grey50") + xlab(paste0("PCoA1 (", var_exp_bray1, "%)")) + ylab(paste0("PCoA2 (", var_exp_bray2, "%)")) + scale_colour_manual(values=common_colors) + scale_shape_manual(values=c(16,17)) + facet_wrap(~facet) + labs(colour="Forest type", shape="Season") + guides(colour=guide_legend(order=1), shape=guide_legend(order=2, override.aes=list(colour="grey60", size=3))) + theme_pubr(border=TRUE) + theme(axis.title=element_text(size=12, face="plain"), strip.background=element_rect(fill="grey20"), strip.text=element_text(color="white", size=14), legend.position=c(0.15,0.25), legend.direction="vertical", legend.key.size=unit(0.35,"cm"), legend.spacing.y=unit(0.15,"cm"), legend.text=element_text(size=8), legend.title=element_text(size=10, face="bold", margin=margin(b=1)), legend.box.margin=margin(2,2,2,2)) + annotate("text", x=-0.1, y=-0.2, size=3.2, label="Monoculture") + annotate("text", x=0.13, y=0.1, size=3.2, label="Restored forests") + annotate("text", x=-0.2, y=0.15, size=3.2, label="Rainforest")
species_composition_total
```

### Phylogenetic composition

```{r}
# Weighted UniFrac distance
wunifrac_dist <- picante::unifrac(total_rel_abundance_t, total_tree)
# Define common color and shape scales
common_colors <- c("RM"= "cyan4", "NR" = "chartreuse3", "RP"= "#7db53eff", "MF" = "forestgreen")
common_shapes <- c("Dry" = 16, "Wet" = 17)
# PCoA
pcoa_w <- cmdscale(wunifrac_dist, eig = TRUE, k = 2)
# Create PCoA dataframe
pcoa_wunifrac_df <- data.frame(Sample=rownames(total_rel_abundance_t), PCoA1=pcoa_w$points[,1], PCoA2=pcoa_w$points[,2], forest_type=env_var$forest_type, season=env_var$season, facet="Total fungi phylogenetic composition")
# Percentage variance explained
var_exp_w1 <- round(pcoa_w$eig[1] / sum(pcoa_w$eig) * 100, 1)
var_exp_w2 <- round(pcoa_w$eig[2] / sum(pcoa_w$eig) * 100, 1)
# Plot Weighted UniFrac PCoA
pcoa_wunifrac_df$forest_type <- factor(pcoa_wunifrac_df$forest_type, levels = c("RM", "NR", "RP", "MF"))
pcoa_wunifrac_df$season <- factor(pcoa_wunifrac_df$season, levels = c("Dry", "Wet"))
# phylogenetic_composition (fixed)
phylogenetic_composition <- ggplot(pcoa_wunifrac_df, aes(x=PCoA1, y=PCoA2, colour=forest_type, shape=season)) + geom_point(size=5, stroke=1) + geom_vline(xintercept=0, linetype="dashed", color="grey50") + geom_hline(yintercept=0, linetype="dashed", color="grey50") + xlab(paste0("PCoA1 (", var_exp_w1, "%)")) + ylab(paste0("PCoA2 (", var_exp_w2, "%)")) + scale_colour_manual(values=common_colors) + scale_shape_manual(values=common_shapes) + facet_wrap(~facet) + labs(colour="Forest type", shape="Season") + guides(colour=guide_legend(order=1), shape=guide_legend(order=2, override.aes=list(colour="grey60", size=3))) + theme_pubr(border=TRUE) + theme(axis.title=element_text(size=12, face="plain"), strip.background=element_rect(fill="grey20"), strip.text=element_text(color="white", size=14), legend.position=c(0.15,0.25), legend.direction="vertical", legend.key.size=unit(0.35,"cm"), legend.spacing.y=unit(0.15,"cm"), legend.text=element_text(size=8), legend.title=element_text(size=10, face="bold", margin=margin(b=1)), legend.box.margin=margin(2,2,2,2)) + annotate("text", x=0.1, y=-0.18, size=3.2, label="Monoculture") + annotate("text", x=0.1, y=0.12, size=3.2, label="Restored forests") + annotate("text", x=-0.28, y=0.05, size=3.2, label="Rainforest")
phylogenetic_composition
```

## AM fungi

### Taxonomic composition

```{r}
AM_rel_abundance_t <- t(AM_rel_abundance_AM)
# Bray-Curtis distance
bray_dist_AM <- vegdist(AM_rel_abundance_t, method = "bray")
# Convert distance to matrix
bray_mat_AM <- as.matrix(bray_dist_AM)
# Total variance (sum of squares)
total_var <- sum(bray_mat_AM^2) / nrow(bray_mat_AM)
# PCoA
bray_pcoa_AM <- cmdscale(bray_dist_AM, eig = TRUE, k = 2)
# PCoA data frame
pcoa_bray_df_AM <- data.frame(Sample=rownames(bray_pcoa_AM$points), PCoA1=bray_pcoa_AM$points[,1], PCoA2=bray_pcoa_AM$points[,2], forest_type=env_var$forest_type, season=env_var$season, facet="AM fungi taxonomic composition")
# Variance explained
var_exp_bray1_AM <- round(bray_pcoa_AM$eig[1] / sum(bray_pcoa_AM$eig) * 100, 1)
var_exp_bray2_AM <- round(bray_pcoa_AM$eig[2] / sum(bray_pcoa_AM$eig) * 100, 1)
# Factor levels
pcoa_bray_df_AM$forest_type <- factor(pcoa_bray_df$forest_type, levels = c("RM", "NR", "RP", "MF"))
pcoa_bray_df_AM$season      <- factor(pcoa_bray_df$season, levels = c("Dry", "Wet"))
species_composition_AM <- ggplot(pcoa_bray_df_AM, aes(x=PCoA1, y=PCoA2, colour=forest_type, shape=season)) + geom_point(size=5, stroke=1) + geom_vline(xintercept=0, linetype="dashed", color="grey50") + geom_hline(yintercept=0, linetype="dashed", color="grey50") + xlab(paste0("PCoA1 (", var_exp_bray1_AM, "%)")) + ylab(paste0("PCoA2 (", var_exp_bray2_AM, "%)")) + scale_colour_manual(values=common_colors) + scale_shape_manual(values=c(16,17)) + facet_wrap(~facet) + labs(colour="Forest type", shape="Season") + guides(colour=guide_legend(order=1), shape=guide_legend(order=2, override.aes=list(colour="grey60", size=3))) + theme_pubr(border=TRUE) + theme(axis.title=element_text(size=12, face="plain"), strip.background=element_rect(fill="grey20"), strip.text=element_text(color="white", size=14), legend.position=c(0.85,0.75), legend.direction="vertical", legend.key.size=unit(0.35,"cm"), legend.spacing.y=unit(0.15,"cm"), legend.text=element_text(size=8), legend.title=element_text(size=10, face="bold", margin=margin(b=1)), legend.box.margin=margin(2,2,2,2)) + annotate("text", x=-0.05, y=0.3, size=3.2, label="Monoculture") + annotate("text", x=-0.2, y=-0.05, size=3.2, label="Restored forests") + annotate("text", x=0.4, y=-0.11, size=3.2, label="Rainforest")
species_composition_AM
```

### Phylogenetic composition

```{r}
# Weighted UniFrac distance
wunifrac_dist_AM <- picante::unifrac(AM_rel_abundance_t, AM_tree)
# PCoA
pcoa_w_AM <- cmdscale(wunifrac_dist_AM, eig = TRUE, k = 2)
# Create PCoA dataframe
pcoa_wunifrac_df_AM <- data.frame(Sample=rownames(AM_rel_abundance_t), PCoA1=pcoa_w_AM$points[,1], PCoA2=pcoa_w_AM$points[,2], forest_type=env_var$forest_type, season=env_var$season, facet="AM fungi phylogenetic composition")
# Percentage variance explained
var_exp_w1_AM <- round(pcoa_w_AM$eig[1] / sum(pcoa_w_AM$eig) * 100, 1)
var_exp_w2_AM <- round(pcoa_w_AM$eig[2] / sum(pcoa_w_AM$eig) * 100, 1)
# Plot Weighted UniFrac PCoA
pcoa_wunifrac_df_AM$forest_type <- factor(pcoa_wunifrac_df_AM$forest_type, levels = c("RM", "NR", "RP", "MF"))
pcoa_wunifrac_df_AM$season <- factor(pcoa_wunifrac_df_AM$season, levels = c("Dry", "Wet"))
phylogenetic_composition_AM <- ggplot(pcoa_wunifrac_df_AM, aes(x=PCoA1, y=PCoA2, colour=forest_type, shape=season)) + geom_point(size=5, stroke=1) + geom_vline(xintercept=0, linetype="dashed", color="grey50") + geom_hline(yintercept=0, linetype="dashed", color="grey50") + xlab(paste0("PCoA1 (", var_exp_w1_AM, "%)")) + ylab(paste0("PCoA2 (", var_exp_w2_AM, "%)")) + scale_colour_manual(values=common_colors) + scale_shape_manual(values=common_shapes) + facet_wrap(~facet) + labs(colour="Forest type", shape="Season") + guides(colour=guide_legend(order=1), shape=guide_legend(order=2, override.aes=list(colour="grey60", size=3))) + theme_pubr(border=TRUE) + theme(axis.title=element_text(size=12, face="plain"), strip.background=element_rect(fill="grey20"), strip.text=element_text(color="white", size=14), legend.position=c(0.6,0.25), legend.direction="vertical", legend.key.size=unit(0.35,"cm"), legend.spacing.y=unit(0.15,"cm"), legend.text=element_text(size=8), legend.title=element_text(size=10, face="bold", margin=margin(b=1)), legend.box.margin=margin(2,2,2,2)) + annotate("text", x=0, y=0.15, size=3.2, label="Monoculture") + annotate("text", x=-0.2, y=0.12, size=3.2, label="Restored forests") + annotate("text", x=0.25, y=-0.05, size=3.2, label="Rainforest")
phylogenetic_composition_AM
```

## Effect of restoration on community composition

### Total fungi

#### Taxonomic composition

```{r}
coords <- location[, c("Longitude", "Latitude")]
coords$Longitude <- as.numeric(coords$Longitude)
coords$Latitude  <- as.numeric(coords$Latitude)
# Distance matrix
coord_dist <- dist(coords)
# PCNM computation
pcnm_obj  <- pcnm(coord_dist)
pcnm_vars <- as.data.frame(pcnm_obj$vectors)
sig_pcnm <- data.frame(PCNM = colnames(pcnm_vars), pvalue = NA)

for (i in seq_len(ncol(pcnm_vars))) {
  tmp_df <- data.frame(pcnm = pcnm_vars[, i])
  test_i <- adonis2(vegdist(total_rel_abundance_t, "bray") ~ pcnm,
                    data = tmp_df, permutations = 9999)
  sig_pcnm$pvalue[i] <- test_i$`Pr(>F)`[1]
}

# Keep PCNMs with p < 0.05
selected_pcnm <- which(sig_pcnm$pvalue < 0.05)
pcnm_selected <- pcnm_vars[, selected_pcnm, drop = FALSE]
env_pcnm <- cbind(env_var, pcnm_selected)
# Set permutation block
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# PERMANOVA with marginal (Type III-like) test
adonis_res <- adonis2(bray_dist_total ~ restoration + season + PCNM1 + PCNM2 + PCNM3 + PCNM4, data=env_pcnm, permutations=999, strata=env_pcnm$Plot_ID, by="margin")
adonis_res
# Tests if group dispersions are equal
disp <- betadisper(bray_dist_total, env_pcnm$restoration)
anova(disp)   
# Define restoration levels
rest_levels <- unique(env_pcnm$restoration)
# Get all pairwise combinations
pairs <- combn(rest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_total)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$restoration %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ restoration + season + PCNM1 + PCNM2 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

```{r}
# Define forest levels
forest_levels <- unique(env_pcnm$forest_types)
# Get all pairwise combinations
pairs <- combn(forest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_total)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$forest_types %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ forest_types + season + PCNM1 + PCNM2 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

#### Phylogenetic composition

```{r}
coords <- location[, c("Longitude", "Latitude")]
coords$Longitude <- as.numeric(coords$Longitude)
coords$Latitude  <- as.numeric(coords$Latitude)
# Distance matrix
coord_dist <- dist(coords)
# PCNM computation
pcnm_obj  <- pcnm(coord_dist)
pcnm_vars <- as.data.frame(pcnm_obj$vectors)
sig_pcnm <- data.frame(PCNM = colnames(pcnm_vars), pvalue = NA)
# Weighted UniFrac distance
wunifrac_dist <- picante::unifrac(total_rel_abundance_t, total_tree)

for (i in seq_len(ncol(pcnm_vars))) {
  tmp_df <- data.frame(pcnm = pcnm_vars[, i])
  test_i <- adonis2(wunifrac_dist ~ pcnm,
                    data = tmp_df, permutations = 9999)
  sig_pcnm$pvalue[i] <- test_i$`Pr(>F)`[1]
}

# Keep PCNMs with p < 0.05
selected_pcnm <- which(sig_pcnm$pvalue < 0.05)
pcnm_selected <- pcnm_vars[, selected_pcnm, drop = FALSE]
env_pcnm <- cbind(env_var, pcnm_selected)

# Set permutation block
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# PERMANOVA with marginal (Type III-like) test
adonis_res <- adonis2(wunifrac_dist ~ restoration + season + PCNM1 + PCNM2 + PCNM3 + PCNM4, data=env_pcnm, permutations=999, strata=env_pcnm$Plot_ID, by="margin")
adonis_res
# Tests if group dispersions are equal
disp <- betadisper(wunifrac_dist, env_pcnm$restoration)
anova(disp)   
# Define restoration levels
rest_levels <- unique(env_pcnm$restoration)
# Get all pairwise combinations
pairs <- combn(rest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
wunifrac_mat <- as.matrix(wunifrac_dist)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$restoration %in% p
  dist_subset <- wunifrac_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ restoration + season + PCNM1 + PCNM2 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

```{r}
# Define forest levels
forest_levels <- unique(env_pcnm$forest_types)
# Get all pairwise combinations
pairs <- combn(forest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
wunifrac_mat <- as.matrix(wunifrac_dist)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$forest_types %in% p
  dist_subset <- wunifrac_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ forest_types + season + PCNM1 + PCNM2 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

### AM fungi

#### Taxonomic composition

```{r}
coords <- location[, c("Longitude", "Latitude")]
coords$Longitude <- as.numeric(coords$Longitude)
coords$Latitude  <- as.numeric(coords$Latitude)
# Distance matrix
coord_dist <- dist(coords)
# PCNM computation
pcnm_obj  <- pcnm(coord_dist)
pcnm_vars <- as.data.frame(pcnm_obj$vectors)
sig_pcnm <- data.frame(PCNM = colnames(pcnm_vars), pvalue = NA)

for (i in seq_len(ncol(pcnm_vars))) {
  tmp_df <- data.frame(pcnm = pcnm_vars[, i])
  test_i <- adonis2(bray_dist_AM ~ pcnm,
                    data = tmp_df, permutations = 9999)
  sig_pcnm$pvalue[i] <- test_i$`Pr(>F)`[1]
}

# Keep PCNMs with p < 0.05
selected_pcnm <- which(sig_pcnm$pvalue < 0.05)
pcnm_selected <- pcnm_vars[, selected_pcnm, drop = FALSE]
env_pcnm <- cbind(env_var, pcnm_selected)
# Set permutation block
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# PERMANOVA with marginal (Type III-like) test
adonis_res <- adonis2(bray_dist_AM ~ restoration + season + PCNM1 + PCNM2 + PCNM3 + PCNM4, data=env_pcnm, permutations=999, strata=env_pcnm$Plot_ID, by="margin")
adonis_res
# Tests if group dispersions are equal
disp <- betadisper(bray_dist_AM, env_pcnm$restoration)
anova(disp)   
# Define restoration levels
rest_levels <- unique(env_pcnm$restoration)
# Get all pairwise combinations
pairs <- combn(rest_levels, 2, simplify = FALSE)

# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_AM)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$restoration %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ restoration + season + PCNM1 + PCNM2 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

```{r}
# Define forest levels
forest_levels <- unique(env_pcnm$forest_types)
# Get all pairwise combinations
pairs <- combn(forest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_AM)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$forest_types %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ forest_types + season + PCNM1 + PCNM2 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

#### Phylogenetic composition

```{r}
coords <- location[, c("Longitude", "Latitude")]
coords$Longitude <- as.numeric(coords$Longitude)
coords$Latitude  <- as.numeric(coords$Latitude)
# Distance matrix
coord_dist <- dist(coords)
# PCNM computation
pcnm_obj  <- pcnm(coord_dist)
pcnm_vars <- as.data.frame(pcnm_obj$vectors)
sig_pcnm <- data.frame(PCNM = colnames(pcnm_vars), pvalue = NA)

for (i in seq_len(ncol(pcnm_vars))) {
  tmp_df <- data.frame(pcnm = pcnm_vars[, i])
  test_i <- adonis2(wunifrac_dist_AM ~ pcnm,
                    data = tmp_df, permutations = 9999)
  sig_pcnm$pvalue[i] <- test_i$`Pr(>F)`[1]
}

# Keep PCNMs with p < 0.05
selected_pcnm <- which(sig_pcnm$pvalue < 0.05)
pcnm_selected <- pcnm_vars[, selected_pcnm, drop = FALSE]

env_pcnm <- cbind(env_var, pcnm_selected)

# Set permutation block
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID

# PERMANOVA with marginal (Type III-like) test
adonis_res <- adonis2(wunifrac_dist_AM ~ restoration + season + PCNM1 + PCNM2 + PCNM3 + PCNM4, data=env_pcnm, permutations=999, strata=env_pcnm$Plot_ID, by="margin")
adonis_res
# tests if group dispersions are equal
disp <- betadisper(wunifrac_dist_AM, env_pcnm$restoration)
anova(disp)   
# Define restoration levels
rest_levels <- unique(env_pcnm$restoration)
# Get all pairwise combinations
pairs <- combn(rest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
unifrac_mat <- as.matrix(wunifrac_dist_AM)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$restoration %in% p
  dist_subset <- unifrac_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ restoration + season + PCNM1 + PCNM2 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

```{r}
# Define forest levels
forest_levels <- unique(env_pcnm$forest_types)
# Get all pairwise combinations
pairs <- combn(forest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
wunifrac_mat <- as.matrix(wunifrac_dist_AM)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$forest_types %in% p
  dist_subset <- wunifrac_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ forest_types + season + PCNM1 + PCNM2 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

# Functional groups relative abundance

## Relative abundance

```{r}
trophic_mode <- read.csv("data/trophic_mode.csv",header = TRUE)

trophic_mode$Forest <- factor (trophic_mode$Forest, levels = c("RM", "NR", "RP", "MF"))


trophic_mode$Category <- factor(trophic_mode$Category, levels = c("Saprophytic fungi", "Symbiotic fungi", "Pathogenic fungi"))


common_colors <- c("RM"= "cyan4", "NR" = "chartreuse3", "RP"= "#7db53eff", "MF" = "forestgreen")


trophic_mode_plot <- ggplot(trophic_mode,
       aes(x = Forest,
           y = Relative_abundance,
           fill = Forest)) +
  geom_boxplot(width = 0.5, outlier.shape = NA)+
  geom_jitter(aes(shape = Season), width = 0.05, size = 1.5, alpha = 0.7) +
  geom_vline(xintercept = 1.6, color = "grey50", linetype = "dashed") +
  geom_vline(xintercept = 3.5, color = "grey50", linetype = "dashed") +
  scale_fill_manual(values = common_colors)+
  facet_wrap(~Category, scales = "free")+
  theme_bw() +
  labs(x = "Forest",
       y = "Relative abundance (%)") +
theme_pubr(border = TRUE, legend = "none")+
  theme(strip.background = element_rect(fill = "grey20"), strip.text = element_text(size = 12, face = "plain", color = "white")) + annotate("text", x = 1, y = Inf, label = "Monoculture", vjust = 1.8, size = 3.2, fontface = "plain") + annotate("text", x = 2.5, y = Inf, label = "Restored forests", vjust = 1.8, size = 3.2, fontface = "plain") + annotate("text", x = 4, y = Inf, label = "Rainforest", vjust = 1.8, size = 3.2, fontface = "plain") + scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

trophic_mode_plot
```

```{r}
saprotroph_category <- read.csv("data/saprotroph_category.csv",header = TRUE)

saprotroph_category$Forest <- factor (saprotroph_category$Forest, levels = c("RM", "NR", "RP", "MF"))


saprotroph_category$Category <- factor(saprotroph_category$Category, levels = c("Soil saprotroph", "Litter saprotroph", "Wood saprotroph"))


common_colors <- c("RM"= "cyan4", "NR" = "chartreuse3", "RP"= "#7db53eff", "MF" = "forestgreen")


saprotroph_plot <- ggplot(saprotroph_category,
       aes(x = Forest,
           y = Relative_abundance,
           fill = Forest)) +
  geom_boxplot(width = 0.5, outlier.shape = NA)+
  geom_jitter(aes(shape = Season), width = 0.05, size = 1.5, alpha = 0.7) +
  geom_vline(xintercept = 1.6, color = "grey50", linetype = "dashed") +
  geom_vline(xintercept = 3.5, color = "grey50", linetype = "dashed") +
  scale_fill_manual(values = common_colors)+
  facet_wrap(~Category, scales = "free")+
  theme_bw() +
  labs(x = "Forest",
       y = "Relative abundance (%)") +
theme_pubr(border = TRUE, legend = "none")+
  theme(strip.background = element_rect(fill = "grey20"), strip.text = element_text(size = 12, face = "plain", color = "white")) + annotate("text", x = 1, y = Inf, label = "Monoculture", vjust = 1.8, size = 3.2, fontface = "plain") + annotate("text", x = 2.5, y = Inf, label = "Restored forests", vjust = 1.8, size = 3.2, fontface = "plain") + annotate("text", x = 4, y = Inf, label = "Rainforest", vjust = 1.8, size = 3.2, fontface = "plain") + scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

saprotroph_plot
```

## Composition

```{r}
# Read data
saprotroph_composition <- read.csv("data/saprotroph_composition.csv",
                                   row.names = 1,
                                   header = TRUE)

saprotroph_rel_abundance <- sweep(
  saprotroph_composition,
  2,
  rep(34300, ncol(saprotroph_composition)),
  FUN = "/"
)

saprotroph_rel_abundance_t <- t(saprotroph_rel_abundance)

bray_dist_sapro <- vegdist(
  saprotroph_rel_abundance_t,
  method = "bray"
)


# Total variance
bray_mat <- as.matrix(bray_dist_sapro)
total_var <- sum(bray_mat^2) / nrow(bray_mat)
total_var

# PCoA
bray_pcoa_sapro <- cmdscale(bray_dist_sapro,
                            eig = TRUE,
                            k = 2)

# PCoA dataframe
pcoa_bray_df <- data.frame(
  Sample = rownames(bray_pcoa_sapro$points),
  PCoA1 = bray_pcoa_sapro$points[,1],
  PCoA2 = bray_pcoa_sapro$points[,2],
  forest_type = env_var$forest_type,
  season = env_var$season,
  facet = "Saprophytic fungi"
)

# Factor levels
pcoa_bray_df$forest_type <- factor(pcoa_bray_df$forest_type,
                                   levels = c("RM", "NR", "RP", "MF"))
pcoa_bray_df$season <- factor(pcoa_bray_df$season,
                              levels = c("Dry", "Wet"))

# Variance explained
var_exp_bray1 <- round(bray_pcoa_sapro$eig[1] /
                         sum(bray_pcoa_sapro$eig) * 100, 1)
var_exp_bray2 <- round(bray_pcoa_sapro$eig[2] /
                         sum(bray_pcoa_sapro$eig) * 100, 1)

# Colors and shapes
common_colors <- c("RM" = "cyan4",
                   "NR" = "chartreuse3",
                   "RP" = "#7db53eff",
                   "MF" = "forestgreen")

common_shapes <- c("Dry" = 16,
                   "Wet" = 17)

# Plot
species_composition_sapro <- ggplot(
  pcoa_bray_df,
  aes(x = PCoA1,
      y = PCoA2,
      colour = forest_type,
      shape = season)
) +
  geom_point(size = 5, stroke = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  xlab(paste0("PCoA1 (", var_exp_bray1, "%)")) +
  ylab(paste0("PCoA2 (", var_exp_bray2, "%)")) +
  scale_colour_manual(values = common_colors) +
  scale_shape_manual(values = common_shapes) +
  facet_wrap(~facet) +
  labs(colour = "Forest type",
       shape = "Season") +
  guides(
    colour = guide_legend(order = 1),
    shape = guide_legend(
      order = 2,
      override.aes = list(colour = "grey60", size = 3)
    )
  ) +
  theme_pubr(border = TRUE) +
  theme(
    axis.title = element_text(size = 12),
    strip.background = element_rect(fill = "grey20"),
    strip.text = element_text(color = "white", size = 14),
    legend.position = c(0.85, 0.25),
    legend.direction = "horizontal",
    legend.key.size = unit(0.35, "cm"),
    legend.spacing.y = unit(0.15, "cm"),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, face = "bold",
                                margin = margin(b = 1)),
    legend.box.margin = margin(2, 2, 2, 2)
  ) +
  annotate("text", x = 0.22, y = -0.2, size = 3.2,
           label = "Monoculture") +
  annotate("text", x = -0.13, y = 0.1, size = 3.2,
           label = "Restored forests") +
  annotate("text", x = 0.28, y = 0.2, size = 3.2,
           label = "Rainforest")

species_composition_sapro




symbiotroph_composition <- read.csv("data/symbiotroph_composition.csv",
                                    row.names = 1,
                                    header = TRUE)

symbiotroph_rel_abundance <- sweep(
  symbiotroph_composition,
  2,
  rep(34300, ncol(symbiotroph_composition)),
  FUN = "/"
)

symbiotroph_rel_abundance_t <- t(symbiotroph_rel_abundance)

bray_dist_symbio <- vegdist(
  symbiotroph_rel_abundance_t,
  method = "bray"
)

# Total variance
bray_mat <- as.matrix(bray_dist_symbio)
total_var <- sum(bray_mat^2) / nrow(bray_mat)
total_var

# PCoA
bray_pcoa_symbio <- cmdscale(bray_dist_symbio,
                             eig = TRUE,
                             k = 2)

# PCoA dataframe
pcoa_bray_df <- data.frame(
  Sample = rownames(bray_pcoa_symbio$points),
  PCoA1 = bray_pcoa_symbio$points[,1],
  PCoA2 = bray_pcoa_symbio$points[,2],
  forest_type = env_var$forest_type,
  season = env_var$season,
  facet = "Symbiotic fungi"
)

# Factor levels
pcoa_bray_df$forest_type <- factor(pcoa_bray_df$forest_type,
                                   levels = c("RM", "NR", "RP", "MF"))
pcoa_bray_df$season <- factor(pcoa_bray_df$season,
                              levels = c("Dry", "Wet"))

# Variance explained
var_exp_bray1 <- round(bray_pcoa_symbio$eig[1] /
                         sum(bray_pcoa_symbio$eig) * 100, 1)
var_exp_bray2 <- round(bray_pcoa_symbio$eig[2] /
                         sum(bray_pcoa_symbio$eig) * 100, 1)

# Colors and shapes
common_colors <- c("RM" = "cyan4",
                   "NR" = "chartreuse3",
                   "RP" = "#7db53eff",
                   "MF" = "forestgreen")

common_shapes <- c("Dry" = 16,
                   "Wet" = 17)

# Plot
species_composition_symbio <- ggplot(
  pcoa_bray_df,
  aes(x = PCoA1,
      y = PCoA2,
      colour = forest_type,
      shape = season)
) +
  geom_point(size = 5, stroke = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  xlab(paste0("PCoA1 (", var_exp_bray1, "%)")) +
  ylab(paste0("PCoA2 (", var_exp_bray2, "%)")) +
  scale_colour_manual(values = common_colors) +
  scale_shape_manual(values = common_shapes) +
  facet_wrap(~facet) +
  labs(colour = "Forest type",
       shape = "Season") +
  guides(
    colour = guide_legend(order = 1),
    shape = guide_legend(
      order = 2,
      override.aes = list(colour = "grey60", size = 3)
    )
  ) +
  theme_pubr(border = TRUE) +
  theme(
    axis.title = element_text(size = 12),
    strip.background = element_rect(fill = "grey20"),
    strip.text = element_text(color = "white", size = 14),
    legend.position = c(0.85, 0.25),
    legend.direction = "horizontal",
    legend.key.size = unit(0.35, "cm"),
    legend.spacing.y = unit(0.15, "cm"),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, face = "bold",
                                margin = margin(b = 1)),
    legend.box.margin = margin(2, 2, 2, 2)
  ) +
  annotate("text", x = -0.22, y = -0.1, size = 3.2,
           label = "Monoculture") +
  annotate("text", x = 0.25, y =-0.15, size = 3.2,
           label = "Restored forests") +
  annotate("text", x = -0.13, y = 0.1, size = 3.2,
           label = "Rainforest")

species_composition_symbio

# Read data
pathotroph_composition <- read.csv("data/pathotroph_composition.csv",
                                   row.names = 1,
                                   header = TRUE)

pathotroph_rel_abundance <- sweep(
  pathotroph_composition,
  2,
  rep(34300, ncol(pathotroph_composition)),
  FUN = "/"
)

pathotroph_rel_abundance_t <- t(pathotroph_rel_abundance)

bray_dist_patho <- vegdist(
  pathotroph_rel_abundance_t,
  method = "bray"
)

# Total variance
bray_mat <- as.matrix(bray_dist_patho)
total_var <- sum(bray_mat^2) / nrow(bray_mat)
total_var

# PCoA
bray_pcoa_patho <- cmdscale(bray_dist_patho,
                            eig = TRUE,
                            k = 2)

# PCoA dataframe
pcoa_bray_df <- data.frame(
  Sample = rownames(bray_pcoa_patho$points),
  PCoA1 = bray_pcoa_patho$points[,1],
  PCoA2 = bray_pcoa_patho$points[,2],
  forest_type = env_var$forest_type,
  season = env_var$season,
  facet = "Pathogenic fungi"
)

# Factor levels
pcoa_bray_df$forest_type <- factor(pcoa_bray_df$forest_type,
                                   levels = c("RM", "NR", "RP", "MF"))
pcoa_bray_df$season <- factor(pcoa_bray_df$season,
                              levels = c("Dry", "Wet"))

# Variance explained
var_exp_bray1 <- round(bray_pcoa_patho$eig[1] /
                         sum(bray_pcoa_patho$eig) * 100, 1)
var_exp_bray2 <- round(bray_pcoa_patho$eig[2] /
                         sum(bray_pcoa_patho$eig) * 100, 1)

# Colors and shapes
common_colors <- c("RM" = "cyan4",
                   "NR" = "chartreuse3",
                   "RP" = "#7db53eff",
                   "MF" = "forestgreen")

common_shapes <- c("Dry" = 16,
                   "Wet" = 17)

# Plot
species_composition_patho <- ggplot(
  pcoa_bray_df,
  aes(x = PCoA1,
      y = PCoA2,
      colour = forest_type,
      shape = season)
) +
  geom_point(size = 5, stroke = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  xlab(paste0("PCoA1 (", var_exp_bray1, "%)")) +
  ylab(paste0("PCoA2 (", var_exp_bray2, "%)")) +
  scale_colour_manual(values = common_colors) +
  scale_shape_manual(values = common_shapes) +
  facet_wrap(~facet) +
  labs(colour = "Forest type",
       shape = "Season") +
  guides(
    colour = guide_legend(order = 1),
    shape = guide_legend(
      order = 2,
      override.aes = list(colour = "grey60", size = 3)
    )
  ) +
  theme_pubr(border = TRUE) +
  theme(
    axis.title = element_text(size = 12),
    strip.background = element_rect(fill = "grey20"),
    strip.text = element_text(color = "white", size = 14),
    legend.position = c(0.15, 0.25),
    legend.direction = "horizontal",
    legend.key.size = unit(0.35, "cm"),
    legend.spacing.y = unit(0.15, "cm"),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, face = "bold",
                                margin = margin(b = 1)),
    legend.box.margin = margin(2, 2, 2, 2)
  ) +
  annotate("text", x = 0.1, y = -0.24, size = 3.2,
           label = "Monoculture") +
  annotate("text", x = 0.1, y = 0.2, size = 3.2,
           label = "Restored forests") +
  annotate("text", x = -0.4, y = 0.1, size = 3.2,
           label = "Rainforest")

species_composition_patho
```

## Effect of restoration on fungal functional group relative abundance

### Saprotrophs

Does restoration enhance relative abundance?

```{r}
# Extract only Saprophytic fungi
saprotroph <- trophic_mode[trophic_mode$Category == "Saprophytic fungi", ]

# To do the same for pathotrophs (adjusting the name to match your exact data)
pathotroph <- trophic_mode[trophic_mode$Category == "Pathogenic fungi", ]

symbiotroph <- trophic_mode[trophic_mode$Category == "Symbiotic fungi", ]

# Set reference level for restoration
saprotroph$restoration <- relevel(factor(saprotroph$restoration), ref = "Monoculture")
# Check normality of the response variable
shapiro.test(saprotroph$Relative_abundance)
# Fit linear mixed-effects model
mod_restoration <- lmer(Relative_abundance ~ restoration + Season + (1 | Plot), data = saprotroph)
# Model summary
summary(mod_restoration)
# Type III ANOVA for fixed effects
anova(mod_restoration, type = 3)
```

Does restoration level rainforest condition?

```{r}
#### Restored vs Reference
# Set reference level for restoration
saprotroph$restoration <- relevel(factor(saprotroph$restoration), ref = "Rainforest")
# Check normality of the response variable
shapiro.test(saprotroph$Relative_abundance)
# Fit linear mixed-effects model
mod_restoration <- lmer(Relative_abundance ~ restoration + Season + (1 | Plot), data = saprotroph)
# Model summary
summary(mod_restoration)
# Type III ANOVA for fixed effects
anova(mod_restoration, type = 3)
```

Are restoration strategies different?

```{r}
# Are restoration strategies different?
saprotroph$Forest <- factor(saprotroph$Forest)
saprotroph$Forest <- relevel(saprotroph$Forest, ref = "NR")
mod_restoration <- lmer(Relative_abundance ~ Forest + Season + (1 | Plot),data = saprotroph)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

### Symbiotrophs

Does restoration enhance relative abundance?

```{r}
# 1. Extract only Symbiotic fungi
symbiotroph <- trophic_mode[trophic_mode$Category == "Symbiotic fungi", ]

# 2. Set reference level for restoration
symbiotroph$restoration <- relevel(factor(symbiotroph$restoration), ref = "Monoculture")

# 3. Check normality of the response variable
shapiro.test(symbiotroph$Relative_abundance)

# 4. Fit linear mixed-effects model
mod_restoration_sym <- lmer(sqrt(Relative_abundance) ~ restoration + Season + (1 | Plot), data = symbiotroph)

sym_resids <- residuals(mod_restoration_sym)
shapiro.test(sym_resids)

# 5. Model summary
summary(mod_restoration_sym)

# 6. Type III ANOVA for fixed effects
anova(mod_restoration_sym, type = 3)
```

Does restoration level rainforest condition?

```{r}
# 1. Extract only Symbiotic fungi
symbiotroph <- trophic_mode[trophic_mode$Category == "Symbiotic fungi", ]

# 2. Set reference level for restoration to Rainforest
symbiotroph$restoration <- relevel(factor(symbiotroph$restoration), ref = "Rainforest")

# 3. Check normality of the response variable
# Correct Syntax
shapiro.test(sqrt(symbiotroph$Relative_abundance))

# 4. Fit linear mixed-effects model
mod_restoration_sym <- lmer(sqrt(Relative_abundance) ~ restoration + Season + (1 | Plot), data = symbiotroph)


sym_resids <- residuals(mod_restoration_sym)
shapiro.test(sym_resids)

# 5. Model summary
summary(mod_restoration_sym)

# 6. Type III ANOVA for fixed effects
anova(mod_restoration_sym, type = 3)

```

Are restoration strategies different?

```{r}
# Set Forest as a factor and change the reference level to NR
symbiotroph$Forest <- factor(symbiotroph$Forest)
symbiotroph$Forest <- relevel(symbiotroph$Forest, ref = "NR")

# Fit the linear mixed-effects model
mod_forest_sym <- lmer(sqrt(Relative_abundance) ~ Forest + Season + (1 | Plot), data = symbiotroph)

# View model coefficients and t-tests
summary(mod_forest_sym)

# View Type III ANOVA table (F-test)
anova(mod_forest_sym, type = 3)
```

### Pathotrophs

Does restoration enhance relative abundance?

```{r}
# 1. Extract only Pathotrophic fungi
pathotroph <- trophic_mode[trophic_mode$Category == "Pathogenic fungi", ]

# 2. Set reference level for restoration
pathotroph$restoration <- relevel(factor(pathotroph$restoration), ref = "Monoculture")

# 3. Check normality of the response variable
shapiro.test(pathotroph$Relative_abundance)

# 4. Fit linear mixed-effects model
mod_restoration_path <- lmer(sqrt(Relative_abundance) ~ restoration + Season + (1 | Plot), data = pathotroph)

# 5. Model summary
summary(mod_restoration_path)

# 6. Type III ANOVA for fixed effects
anova(mod_restoration_path, type = 3)
```

Does restoration level rainforest condition?

```{r}

# 2. Set reference level for restoration to Rainforest
pathotroph$restoration <- relevel(factor(pathotroph$restoration), ref = "Rainforest")


# 4. Fit linear mixed-effects model
mod_restoration_path <- lmer(sqrt(Relative_abundance) ~ restoration + Season + (1 | Plot), data = pathotroph)

# 5. Model summary
summary(mod_restoration_path)

# 6. Type III ANOVA for fixed effects
anova(mod_restoration_path, type = 3)
```

Are restoration strategies different?

```{r}
# Set Forest as a factor and change the reference level to NR
pathotroph$Forest <- factor(pathotroph$Forest)
pathotroph$Forest <- relevel(pathotroph$Forest, ref = "NR")

# Fit the linear mixed-effects model
mod_forest_path <- lmer(sqrt(Relative_abundance) ~ Forest + Season + (1 | Plot), data = pathotroph)

# View model coefficients and t-tests
summary(mod_forest_path)

# View Type III ANOVA table (F-test)
anova(mod_forest_path, type = 3)
```

### Saprotroph categories

```{r}
saprotroph_category <- read.csv("data/saprotroph_category.csv",header = TRUE)

litter_saprotroph <- trophic_mode[saprotroph_category$Category == "Litter saprotroph", ]
soil_saprotroph <- trophic_mode[saprotroph_category$Category == "Soil saprotroph", ]
wood_saprotroph <- trophic_mode[saprotroph_category$Category == "Wood saprotroph", ]

```

```{r}
# Restored forest vs monoculture
# Set reference level for restoration
litter_saprotroph$restoration <- relevel(factor(litter_saprotroph$restoration), ref = "Monoculture")
# Check normality of the response variable
shapiro.test(litter_saprotroph$Relative_abundance)
# Fit linear mixed-effects model
mod_restoration <- lmer(Relative_abundance ~ restoration + Season + (1 | Plot), data = litter_saprotroph)
# Model summary
summary(mod_restoration)
# Type III ANOVA for fixed effects
anova(mod_restoration, type = 3)

```

```{r}
# Set reference level for restoration
soil_saprotroph$restoration <- relevel(factor(soil_saprotroph$restoration), ref = "Monoculture")

# Check normality of the response variable
shapiro.test(soil_saprotroph$Relative_abundance)

# Fit linear mixed-effects model
mod_restoration <- lmer(Relative_abundance ~ restoration + Season + (1 | Plot),
                        data = soil_saprotroph)

# Model summary
summary(mod_restoration)

# Type III ANOVA for fixed effects
anova(mod_restoration, type = 3)


#### Restored vs Reference
# Set reference level for restoration
soil_saprotroph$restoration <- relevel(factor(soil_saprotroph$restoration), ref = "Rainforest")
# Check normality of the response variable
shapiro.test(soil_saprotroph$Relative_abundance)
# Fit linear mixed-effects model
mod_restoration <- lmer(Relative_abundance ~ restoration + Season + (1 | Plot), data = soil_saprotroph)
# Model summary
summary(mod_restoration)
# Type III ANOVA for fixed effects
anova(mod_restoration, type = 3)

# Are restoration strategies different?
soil_saprotroph$Forest <- factor(soil_saprotroph$Forest)
soil_saprotroph$Forest <- relevel(soil_saprotroph$Forest, ref = "NR")
mod_restoration <- lmer(Relative_abundance ~ Forest + Season + (1 | Plot),data = soil_saprotroph)
summary(mod_restoration)
anova(mod_restoration, type = 3)
```

```{r}

# Set reference level for restoration
wood_saprotroph$restoration <- relevel(factor(wood_saprotroph$restoration), ref = "Monoculture")

# Check normality of the response variable
shapiro.test(wood_saprotroph$Relative_abundance)

# Fit linear mixed-effects model
mod_restoration <- lmer(Relative_abundance ~ restoration + Season + (1 | Plot),
                        data = wood_saprotroph)

# Model summary
summary(mod_restoration)

# Type III ANOVA for fixed effects
anova(mod_restoration, type = 3)


# Restored vs Reference

# Set reference level for restoration
wood_saprotroph$restoration <- relevel(factor(wood_saprotroph$restoration), ref = "Rainforest")

# Check normality of the response variable
shapiro.test(wood_saprotroph$Relative_abundance)

# Fit linear mixed-effects model
mod_restoration <- lmer(sqrt(Relative_abundance) ~ restoration + Season + (1 | Plot),
                        data = wood_saprotroph)

# Model summary
summary(mod_restoration)

# Type III ANOVA for fixed effects
anova(mod_restoration, type = 3)


# Are restoration strategies different?

wood_saprotroph$Forest <- factor(wood_saprotroph$Forest)
wood_saprotroph$Forest <- relevel(wood_saprotroph$Forest, ref = "NR")

mod_restoration <- lmer(Relative_abundance ~ Forest + Season + (1 | Plot),
                        data = wood_saprotroph)

summary(mod_restoration)

anova(mod_restoration, type = 3)
```

```{r}
# Restored vs Reference

# Set reference level for restoration
litter_saprotroph$restoration <- relevel(factor(litter_saprotroph$restoration), ref = "Rainforest")

# Check normality of the response variable
shapiro.test(litter_saprotroph$Relative_abundance)

# Fit linear mixed-effects model
mod_restoration <- lmer(Relative_abundance ~ restoration + Season + (1 | Plot),
                        data = litter_saprotroph)

# Model summary
summary(mod_restoration)

# Type III ANOVA for fixed effects
anova(mod_restoration, type = 3)


# Are restoration strategies different?

litter_saprotroph$Forest <- factor(litter_saprotroph$Forest)
litter_saprotroph$Forest <- relevel(litter_saprotroph$Forest, ref = "NR")

mod_restoration <- lmer(Relative_abundance ~ Forest + Season + (1 | Plot),
                        data = litter_saprotroph)

summary(mod_restoration)

anova(mod_restoration, type = 3)
```

## Effect of restoration on fungal functional group composition

### Saprotrophs

```{r}
coords <- location[, c("Longitude", "Latitude")]
coords$Longitude <- as.numeric(coords$Longitude)
coords$Latitude  <- as.numeric(coords$Latitude)
# Distance matrix
coord_dist <- dist(coords)
# PCNM computation
pcnm_obj  <- pcnm(coord_dist)
pcnm_vars <- as.data.frame(pcnm_obj$vectors)
sig_pcnm <- data.frame(PCNM = colnames(pcnm_vars), pvalue = NA)

for (i in seq_len(ncol(pcnm_vars))) {
  tmp_df <- data.frame(pcnm = pcnm_vars[, i])
  test_i <- adonis2(vegdist(saprotroph_rel_abundance_t, "bray") ~ pcnm,
                    data = tmp_df, permutations = 9999)
  sig_pcnm$pvalue[i] <- test_i$`Pr(>F)`[1]
}

bray_dist_sapro <- vegdist(
  saprotroph_rel_abundance_t,
  method = "bray"
)

# Keep PCNMs with p < 0.05
selected_pcnm <- which(sig_pcnm$pvalue < 0.05)
pcnm_selected <- pcnm_vars[, selected_pcnm, drop = FALSE]
env_pcnm <- cbind(env_var, pcnm_selected)
# Set permutation block
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# PERMANOVA with marginal (Type III-like) test
adonis_res <- adonis2(bray_dist_sapro ~ restoration + season + PCNM1 + PCNM3, data=env_pcnm, permutations=999, strata=env_pcnm$Plot_ID, by="margin")
adonis_res
# Tests if group dispersions are equal
disp <- betadisper(bray_dist_sapro, env_pcnm$restoration)
anova(disp)   
# Define restoration levels
rest_levels <- unique(env_pcnm$restoration)
# Get all pairwise combinations
pairs <- combn(rest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_sapro)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$restoration %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ restoration + season + PCNM1 + PCNM3,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

```{r}
# Define forest levels
forest_levels <- unique(env_pcnm$forest_types)
# Get all pairwise combinations
pairs <- combn(forest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_sapro)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$forest_types %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ forest_types + season + PCNM1 + PCNM3,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

### Symbiotroph

```{r}
coords <- location[, c("Longitude", "Latitude")]
coords$Longitude <- as.numeric(coords$Longitude)
coords$Latitude  <- as.numeric(coords$Latitude)
# Distance matrix
coord_dist <- dist(coords)
# PCNM computation
pcnm_obj  <- pcnm(coord_dist)
pcnm_vars <- as.data.frame(pcnm_obj$vectors)
sig_pcnm <- data.frame(PCNM = colnames(pcnm_vars), pvalue = NA)

for (i in seq_len(ncol(pcnm_vars))) {
  tmp_df <- data.frame(pcnm = pcnm_vars[, i])
  test_i <- adonis2(vegdist(symbiotroph_rel_abundance_t, "bray") ~ pcnm,
                    data = tmp_df, permutations = 9999)
  sig_pcnm$pvalue[i] <- test_i$`Pr(>F)`[1]
}

bray_dist_symbio <- vegdist(
  symbiotroph_rel_abundance_t,
  method = "bray"
)

# Keep PCNMs with p < 0.05
selected_pcnm <- which(sig_pcnm$pvalue < 0.05)
pcnm_selected <- pcnm_vars[, selected_pcnm, drop = FALSE]
env_pcnm <- cbind(env_var, pcnm_selected)
# Set permutation block
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# PERMANOVA with marginal (Type III-like) test
adonis_res <- adonis2(bray_dist_symbio ~ restoration + season + PCNM1 + PCNM3 + PCNM4, data=env_pcnm, permutations=999, strata=env_pcnm$Plot_ID, by="margin")
adonis_res
# Tests if group dispersions are equal
disp <- betadisper(bray_dist_symbio, env_pcnm$restoration)
anova(disp)   
# Define restoration levels
rest_levels <- unique(env_pcnm$restoration)
# Get all pairwise combinations
pairs <- combn(rest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_symbio)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$restoration %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ restoration + season + PCNM1 + PCNM3 + PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

```{r}
# Define forest levels
forest_levels <- unique(env_pcnm$forest_types)
# Get all pairwise combinations
pairs <- combn(forest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_symbio)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$forest_types %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ forest_types + season + PCNM1 + PCNM3 +PCNM4,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

### Pathotroph

```{r}
coords <- location[, c("Longitude", "Latitude")]
coords$Longitude <- as.numeric(coords$Longitude)
coords$Latitude  <- as.numeric(coords$Latitude)
# Distance matrix
coord_dist <- dist(coords)
# PCNM computation
pcnm_obj  <- pcnm(coord_dist)
pcnm_vars <- as.data.frame(pcnm_obj$vectors)
sig_pcnm <- data.frame(PCNM = colnames(pcnm_vars), pvalue = NA)

for (i in seq_len(ncol(pcnm_vars))) {
  tmp_df <- data.frame(pcnm = pcnm_vars[, i])
  test_i <- adonis2(vegdist(pathotroph_rel_abundance_t, "bray") ~ pcnm,
                    data = tmp_df, permutations = 9999)
  sig_pcnm$pvalue[i] <- test_i$`Pr(>F)`[1]
}

bray_dist_patho <- vegdist(
  pathotroph_rel_abundance_t,
  method = "bray"
)

# Keep PCNMs with p < 0.05
selected_pcnm <- which(sig_pcnm$pvalue < 0.05)
pcnm_selected <- pcnm_vars[, selected_pcnm, drop = FALSE]
env_pcnm <- cbind(env_var, pcnm_selected)
# Set permutation block
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# PERMANOVA with marginal (Type III-like) test
adonis_res <- adonis2(bray_dist_patho ~ restoration + season + PCNM1 + PCNM2 + PCNM3, data=env_pcnm, permutations=999, strata=env_pcnm$Plot_ID, by="margin")
adonis_res
# Tests if group dispersions are equal
disp <- betadisper(bray_dist_patho, env_pcnm$restoration)
anova(disp)   
# Define restoration levels
rest_levels <- unique(env_pcnm$restoration)
# Get all pairwise combinations
pairs <- combn(rest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_patho)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$restoration %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ restoration + season + PCNM1 + PCNM2 + PCNM3,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

```{r}
# Define forest levels
forest_levels <- unique(env_pcnm$forest_types)
# Get all pairwise combinations
pairs <- combn(forest_levels, 2, simplify = FALSE)
# Convert dist to square matrix
bray_mat <- as.matrix(bray_dist_patho)

for(p in pairs){
  cat("\n=== Pairwise PERMANOVA:", p[1], "vs", p[2], "===\n")
  
  # Subset rows/columns for the two groups
  idx <- env_pcnm$forest_types %in% p
  dist_subset <- bray_mat[idx, idx]
  meta_subset <- env_pcnm[idx, ]
  
  # Convert back to dist object
  dist_subset <- as.dist(dist_subset)
  
  # PERMANOVA
  res <- adonis2(
    dist_subset ~ forest_types + season + PCNM1 + PCNM2 +PCNM3,
    data = meta_subset,
    permutations = 999,
    strata = meta_subset$Plot_ID,
    by = "margin"
  )
  
  print(res)
}
```

## Effect of soil abiotic properties

### Alpha diversity

#### Total fungi

Do soil C, N, P and pH affect observed richness?

```{r}
set.seed(123)
mod_soil_main <- lmer(Richness ~ TC + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)  # Type III ANOVA for fixed effects
```

```{r}
mod_soil_main <- lmer(Richness ~ TN + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)  # Type III ANOVA for fixed effects
```

```{r}
mod_soil_main <- lmer(Richness ~ TP + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)  # Type III ANOVA for fixed effects
```

```{r}
mod_soil_main <- lmer(Richness ~ pH + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)  # Type III ANOVA for fixed effects
```

Do soil C, N, P and pH affect Shannon diversity?

```{r}
mod_soil_main <- lmer(Shannon ~ TC + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(Shannon ~ TN + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(Shannon ~ TP + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(Shannon ~ pH + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

Do soil C, N, P and pH affect PD?

```{r}
mod_soil_main <- lmer(PD ~ TC + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(PD ~ TN + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(PD ~ TP + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(PD ~ pH + forest_types + season + (1 | Plot_ID), data = alpha_total)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

#### AM fungi

Do soil C, N, P and pH affect observed richness?

```{r}
set.seed(123)
mod_soil_main <- lmer(Richness ~ TC + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)  # Type III ANOVA for fixed effects
```

```{r}
mod_soil_main <- lmer(Richness ~ TN + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)  # Type III ANOVA for fixed effects
```

```{r}
mod_soil_main <- lmer(Richness ~ TP + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)  # Type III ANOVA for fixed effects
```

```{r}
mod_soil_main <- lmer(Richness ~ pH + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)  # Type III ANOVA for fixed effects
```

Do soil C, N, P and pH affect Shannon diversity?

```{r}
mod_soil_main <- lmer(Shannon ~ TC + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(Shannon ~ TN + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(Shannon ~ TP + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(Shannon ~ pH + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

Do soil C, N, P and pH affect PD?

```{r}
mod_soil_main <- lmer(PD ~ TC + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(PD ~ TN + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(PD ~ TP + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

```{r}
mod_soil_main <- lmer(PD ~ pH + forest_types + season + (1 | Plot_ID), data = alpha_AM)
summary(mod_soil_main)
anova(mod_soil_main, type = 3)
```

## Composition

### Total fungi

#### Taxonomic composition

```{r}
set.seed(123)
pdbRDA <- dbrda(bray_dist_total ~ TC + TN + TP + pH + Condition(forest_types + season + PCNM1 + PCNM2 + PCNM3),data = env_pcnm)
summary(pdbRDA)
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# Overall soil effect
anova(pdbRDA, permutations = perm)
anova(pdbRDA, by = "margin", permutations = perm)
```

#### Phylogenetic composition

```{r}
set.seed(123)
pdbRDA <- dbrda(wunifrac_dist ~ TC + TN + TP + pH + Condition(forest_types + season + PCNM1 + PCNM2 + PCNM3), data = env_pcnm)
summary(pdbRDA)
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# Overall soil effect
anova(pdbRDA, permutations = perm)
anova(pdbRDA, by = "margin", permutations = perm)
```

### AM fungi

#### Taxonomic composition

```{r}
set.seed(123)
pdbRDA <- dbrda(bray_dist_AM ~ TC + TN + TP + pH + Condition(forest_types + season + PCNM1 + PCNM2 + PCNM3), data = env_pcnm)
summary(pdbRDA)
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# Overall soil effect
anova(pdbRDA, permutations = perm)
anova(pdbRDA, by = "margin", permutations = perm)
```

#### Phylogenetic composition

```{r}
set.seed(123)
pdbRDA <- dbrda(wunifrac_dist_AM ~ TC + TN + TP + pH + Condition(forest_types + season + PCNM1 + PCNM2 + PCNM3),data = env_pcnm)
summary(pdbRDA)
perm <- how(nperm = 999)
setBlocks(perm) <- env_pcnm$Plot_ID
# Overall soil effect
anova(pdbRDA, permutations = perm)
anova(pdbRDA, by = "margin", permutations = perm)
```

# Session info

```{r}
sessionInfo()
```
